`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-04-18 manuscript/manuscript.qmd
Add author
Finalise figures for TIFF export

    email: me@me.com

@@

@@

Qualitative data collection consisted of 25 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021, spread around the hospital’s various wards and departments. Interviews were
semi-structured, and included questions on hospital waste management
practices, perceived best practices, and workplace health, safety, and
risk. The first round of interviews included specific questions on air quality, while the second round focused on waste management more broadly. However, the semi-structured nature of the interviews allowed for
the participants to raise topics of interest, and for unexpected themes
to emerge. These interviews are supplemented by participatory
observation recorded within detailed field notes, and supported by
nearly a decade of sustained involvement and activism by the authors in
the waste management practices of the hospital.

@@

Raw data and analysis-ready processed data is available as an R package at @R-qechairquality. The data underlying the tables and figures of this manuscript are contained in a repository alongside a reproducible document that contains the analysis code and the written narrative of the manuscript [TODO: manuscript software citation].

@@

ggsave(filename = "manuscript/figs/fig-03-double-column.tiff",
       plot = fig03,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

library(ggplot2)

# create sample data
df <- data.frame(x = 1:10, y = 1:10)

# create plot
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  xlab(expression(c('PM'[10], 'PM'[2.5])))

@@

ggsave(filename = "manuscript/figs/fig-04-double-column.tiff",
       plot = fig04,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

    location == "Administration" 
    & indicator == "pm25"
    & air_quality_who_annual == "Interim target 3" ~ exposure_percent + 1,

    location == "uMoyo" 
    & indicator == "pm25"
    & air_quality_who_annual == "Interim target 4" ~ exposure_percent - 1,

@@

  #facet_wrap(~location, ncol = 2) +
  theme_minimal(base_size = 14) +
  # help from: https://stackoverflow.com/questions/34749636/left-justify-text-from-multi-line-facet-labels
  theme(strip.text.x = element_text(hjust = 0),
        strip.text.y = element_text(vjust = 0, size = 4))

@@

@@

> Personally, I have never experienced \[eye discomfort\] because I get
> vaccinated and I am protected including other staff. However, we
> realize that the air can badly affect other people and patients who
> come to this place.

@@

The air quality, as a result of open waste burning at Queen’s is poor and not suitable for a city, let alone a hospital. Over the course of this two-month study, the hazardous limits for both PM parameters were exceeded at all locations except for the Lions Sight hospital. The WHO limits were exceeded fewer than 50 times at five of the locations, and only the monitors at the Lighthouse Clinic and the Guardian Shelter recorded more than 50 instances above the most hazardous limit. The composition of emissions recorded ranged significantly, analysed as the ratio PM~10~:PM~2.5~ likely reflects a range of burning styles and/or fuel (waste) type, and no discernible trend was observed..  Among interview respondents, there was a general  consensus that air quality caused by the burning of waste within the hospital was a problem, although there were significant differences between respondents over their understandings of potential impacts, the effectiveness of various coping mechanisms, and their problematisations linked to the burning of specific waste materials. For most, going home or leaving the Queen’s campus was the only pathway to relief, though for patients who can not leave, endurance was the only option. This work was originally planned as a before-and-after study:  we envisioned reporting these baseline results compared to follow-up measurements once the long-promised incinerator was commissioned.  However, nearly 4 years later, the incinerator is rarely functional (partly for technical reasons, partly for financial ones) and the smoke hovering over Queen’s is still thick.  Waste management has long been known as an environmental health issue, but in the case of open trash burning at a medical facility, it becomes more:  for some of the most vulnerable, especially those with TB, asthma or other chronic disease, it could be a matter of life or death.

@@

dat_in |> 
  ggplot(aes(x = date_time, y = value)) +
  geom_line(alpha = 0.5) +
  labs(title = "Hospital Air Quality",
       subtitle = "PM~2.5~ and PM~10~ measurements in 5-min intervals at eight locations") +
  ylab(bquote(uq/m^3)) +
  facet_grid(indicator~location) +
  theme_bw(base_size = 14) + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

@@

1.  Raspberry Pi 3 Model B single-board computer technical
specifications:
-   Broadcom BCM2837 64bit Quad Core Processor powered Single Board
Computer running at 1.2GHz 1GB RAM
-   BCM43438 WiFi on board
-   Bluetooth Low Energy (BLE) on board
-   40pin extended GPIO
-   4 x USB 2 ports 4 pole Stereo output and Composite video port
-   Full size HDMI CSI camera port for connecting the Raspberry Pi
camera
-   DSI display port for connecting the Raspberry Pi touch screen
display
-   Micro SD port for loading your operating system and storing data
-   Upgraded switched Micro USB power source (now supports up to 2.4
Amps)
-   Same form factor as the Pi 2 Model B, however the LEDs have
changed position
-   RPI3-MODB-16GB-NOOBS
2.  Nova PM SDS011 Sensor
-   See Data Sheet attachment for technical specifications
3.  MCP3008 microchip
-   analogue-to-digital converter (ADC)
4.  Romoss Sense8+ 30000mAh QC Type-C
-   Power bank battery
5.  Pipe
-   clear rubber pipe attainable from any local hardware store
6.  Box
-   any box that can fit all the components comfortably can be used.
The box will need to have a whole drilled in it for the pipe.

### Software ‘components’ and data management {.unnumbered}

1. The Raspberry Pi operating system runs off of Rasbian software on a NOOBS scandisk. 
2. Air quality monitoring code is written in Python using the [aqi library](https://pypi.org/project/python-aqi/) (A library of algorithms to convert between AQI value and pollutant concentration). The code is stored on the [Open Data Durban GitHub account](https://github.com/opendatadurban/hospital_stations) and is free for all to access. To be able to collect data, the device needs to be connected to a reliable wifi network, the SSID and password need to be coded onto the device during step 2 in the building method.
3. Once the whole device has been put together and placed on site, to access that data we use a combination of programs:
- [Putty](https://www.putty.org/) - this is an SSH client that we use to gain access to the Raspberry Pi from another computer. To be able to access the Raspberry Pi in question, the computer in use and the air quality monitoring device need to be on the same network for the connection to be successful. 
- [Winscp](https://winscp.net/eng/download.php) - this is an FTP client that we use, in conjunction with Putty, to be able to transfer data files from the air quality monitoring device to one’s computer.
- [Remote.it](https://app.remote.it/) - this is an application that supports remote SSH, so this is used when wanting to access data from a device that is not connected to the same network as the computer in use. Remote.it is used together with Putty and Winscp to access the air quality data remotely.

@@

1. Set up the Raspberry Pi with the Rasbian software using the NOOBS scandisk. (insert the scandisk into the Raspberry Pi) A comprehensive start up guide can be found here. 
2. Next we load the code onto the Raspberry Pi. The code can be found on the Open Data Durban Hospital Stations Repository Github. To use it you can follow the steps in the file called Initiate_WS.txt. Once all the steps have been completed, the Raspberry Pi is now a device that can be connected to a particulate matter monitoring sensor.  
3. Connect the hardware components
- Connect the Raspberry Pi to the Nova PM Sensor using the break out board that comes with the Sensor. 
- Connect the battery to the Raspberry Pi via the micro port
- Attach the rubber pipe to the sensor 
- Assemble all components inside the box, with the pipe coming out of the drilled hole
- To prevent dirt entering the system, close off all openings between the pipe and the box with silicone putty
- Secure box in an elevated position (not too close to the ground) and with the pipe facing the direction of the inflow of the air under investigation.

>>>>

    orcid: 0009-0004-1389-5541
    email: heiko.heilgendorff@gmail.com

@@

  freeze: true

@@

Qualitative data collection consisted of 25 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021, spread around the hospital's
various wards and departments. Interviews were semi-structured, and
included questions on hospital waste management practices, perceived
best practices, and workplace health, safety, and risk. The first round
of interviews included specific questions on air quality, while the
second round focused on waste management more broadly. However, the
semi-structured nature of the interviews allowed for the participants to
raise topics of interest, and for unexpected themes to emerge. These
interviews are supplemented by participatory observation recorded within
detailed field notes, and supported by nearly a decade of sustained
involvement and activism by the authors in the waste management
practices of the hospital.

@@

Raw data and analysis-ready processed data is available as an R package
at @R-qechairquality. The data underlying the tables and figures of this
manuscript are contained in a repository alongside a reproducible
document that contains the analysis code and the written narrative of
the manuscript \[TODO: manuscript software citation\].

@@

fig03

ggsave(filename = here::here("manuscript/figs/fig-03-double-column.tiff"),
       plot = fig03,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

@@

ggsave(filename = here::here("manuscript/figs/fig-04-double-column.tiff"),
       plot = fig04,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

    location == "Administration" 
    & indicator == "pm2.5"
    & air_quality_who_annual == "Interim target 3" ~ exposure_percent + 1,

    location == "Guardian Shelter" 
    & indicator == "pm10"
    & air_quality_who_annual == "Interim target 4" ~ exposure_percent + 1,

    location == "uMoyo" 
    & indicator == "pm2.5"
    & air_quality_who_annual == "Interim target 4" ~ exposure_percent - 1,

@@

  #facet_wrap(~location, ncol = 2) +
  #theme_minimal(base_size = 14) +
  # help from: https://stackoverflow.com/questions/34749636/left-justify-text-from-multi-line-facet-labels
  theme(strip.text.x = element_text(hjust = 0),
        strip.text.y = element_text(vjust = 0, size = 4),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 4)) #+

@@

ggsave(filename = here::here("manuscript/figs/fig-05-double-column.tiff"),
       plot = fig05,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

> Personally, I have never experienced \[eye discomfort\] because I get
> vaccinated and I am protected including other staff. However, we
> realise that the air can badly affect other people and patients who
> come to this place.

@@

The air quality, as a result of open waste burning at Queen's is poor
and not suitable for a city, let alone a hospital. Over the course of
this two-month study, the hazardous limits for both PM parameters were
exceeded at all locations except for the Lions Sight hospital. The WHO
limits were exceeded fewer than 50 times at five of the locations, and
only the monitors at the Lighthouse Clinic and the Guardian Shelter
recorded more than 50 instances above the most hazardous limit. The
composition of emissions recorded ranged significantly, analysed as the
ratio PM~10~:PM~2.5~ likely reflects a range of burning styles and/or
fuel (waste) type, and no discernible trend was observed.. Among
interview respondents, there was a general consensus that air quality
caused by the burning of waste within the hospital was a problem,
although there were significant differences between respondents over
their understandings of potential impacts, the effectiveness of various
coping mechanisms, and their problematisations linked to the burning of
specific waste materials. For most, going home or leaving the Queen's
campus was the only pathway to relief, though for patients who can not
leave, endurance was the only option. This work was originally planned
as a before-and-after study: we envisioned reporting these baseline
results compared to follow-up measurements once the long-promised
incinerator was commissioned. However, nearly 4 years later, the
incinerator is rarely functional (partly for technical reasons, partly
for financial ones) and the smoke hovering over Queen's is still thick.
Waste management has long been known as an environmental health issue,
but in the case of open trash burning at a medical facility, it becomes
more: for some of the most vulnerable, especially those with TB, asthma
or other chronic disease, it could be a matter of life or death.

@@

figA1 <- dat_in |> 
  ggplot(aes(x = date_time, y = value)) +
  geom_line(alpha = 0.5) +
  #labs(title = "Hospital Air Quality",
  #     subtitle = "PM~2.5~ and PM~10~ measurements in 5-min intervals at eight locations") +
  ylab(bquote(uq/m^3)) +
  labs(x = NULL) +
  facet_grid(indicator~location) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figA1

ggsave(filename = here::here("manuscript/figs/fig-A1-double-column.tiff"),
       plot = figA1,
       dpi = param_dpi,
       width = param_width_double,
       height = param_width_double * width_height_ratio,
       units = "mm")

@@

1.  Raspberry Pi 3 Model B single-board computer technical
    specifications:

-   Broadcom BCM2837 64bit Quad Core Processor powered Single Board
    Computer running at 1.2GHz 1GB RAM
-   BCM43438 WiFi on board
-   Bluetooth Low Energy (BLE) on board
-   40pin extended GPIO
-   4 x USB 2 ports 4 pole Stereo output and Composite video port
-   Full size HDMI CSI camera port for connecting the Raspberry Pi
    camera
-   DSI display port for connecting the Raspberry Pi touch screen
    display
-   Micro SD port for loading your operating system and storing data
-   Upgraded switched Micro USB power source (now supports up to 2.4
    Amps)
-   Same form factor as the Pi 2 Model B, however the LEDs have changed
    position
-   RPI3-MODB-16GB-NOOBS

2.  Nova PM SDS011 Sensor

-   See Data Sheet attachment for technical specifications

3.  MCP3008 microchip

-   analogue-to-digital converter (ADC)

4.  Romoss Sense8+ 30000mAh QC Type-C

-   Power bank battery

5.  Pipe

-   clear rubber pipe attainable from any local hardware store

6.  Box

-   any box that can fit all the components comfortably can be used. The
    box will need to have a whole drilled in it for the pipe.

### Software 'components' and data management {.unnumbered}

1.  The Raspberry Pi operating system runs off of Rasbian software on a
    NOOBS scandisk.
2.  Air quality monitoring code is written in Python using the [aqi
    library](https://pypi.org/project/python-aqi/) (A library of
    algorithms to convert between AQI value and pollutant
    concentration). The code is stored on the [Open Data Durban GitHub
    account](https://github.com/opendatadurban/hospital_stations) and is
    free for all to access. To be able to collect data, the device needs
    to be connected to a reliable wifi network, the SSID and password
    need to be coded onto the device during step 2 in the building
    method.
3.  Once the whole device has been put together and placed on site, to
    access that data we use a combination of programs:

-   [Putty](https://www.putty.org/) - this is an SSH client that we use
    to gain access to the Raspberry Pi from another computer. To be able
    to access the Raspberry Pi in question, the computer in use and the
    air quality monitoring device need to be on the same network for the
    connection to be successful.
-   [Winscp](https://winscp.net/eng/download.php) - this is an FTP
    client that we use, in conjunction with Putty, to be able to
    transfer data files from the air quality monitoring device to one's
    computer.
-   [Remote.it](https://app.remote.it/) - this is an application that
    supports remote SSH, so this is used when wanting to access data
    from a device that is not connected to the same network as the
    computer in use. Remote.it is used together with Putty and Winscp to
    access the air quality data remotely.

@@

1.  Set up the Raspberry Pi with the Rasbian software using the NOOBS
    scandisk. (insert the scandisk into the Raspberry Pi) A
    comprehensive start up guide can be found here.
2.  Next we load the code onto the Raspberry Pi. The code can be found
    on the Open Data Durban Hospital Stations Repository Github. To use
    it you can follow the steps in the file called Initiate_WS.txt. Once
    all the steps have been completed, the Raspberry Pi is now a device
    that can be connected to a particulate matter monitoring sensor.\
3.  Connect the hardware components

-   Connect the Raspberry Pi to the Nova PM Sensor using the break out
    board that comes with the Sensor.
-   Connect the battery to the Raspberry Pi via the micro port
-   Attach the rubber pipe to the sensor
-   Assemble all components inside the box, with the pipe coming out of
    the drilled hole
-   To prevent dirt entering the system, close off all openings between
    the pipe and the box with silicone putty
-   Secure box in an elevated position (not too close to the ground) and
    with the pipe facing the direction of the inflow of the air under
    investigation.

### 2023-04-18 manuscript/manuscript.tex
Finalise figures for TIFF export

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangerlo@gmail.comm} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[1]{Saloni Vijay%
%
}
 \ead{svijay@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{me@me.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10 µm (PM\textsubscript{10})
and less than 2.5 µm (PM\textsubscript{2.5}), expressed as the mass of
PM per volume of air (µg PMx/m\textsuperscript{3} air) were recorded
every 5 minutes at 8 locations across the QECH for 2 months. Qualitative
data collection consisted of 57 interviews with patients, caregivers and
hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators). Our results show that safe air quality
thresholds are consistently exceeded across space and time and that the
most problematic air quality surrounds the shelter for caregivers and
those receiving treatment for HIV/AIDs. Moreover, staff and visitors are
severely impacted by the poor air quality within the space, but feel
powerless to make changes or address complaints. Waste management
interventions are desperately needed lest the patients who arrive at
Queen's leave with more health issues than the ones with which they
arrived.
\end{abstract}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, boxrule=0pt, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

Qualitative data collection consisted of 26 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021. Interviews were
semi-structured, and included questions on hospital waste management
practices, perceived best practices, and workplace health, safety, and
risk. However, the semi-structured nature of the interviews allowed for
the participants to raise topics of interest, and for unexpected themes
to emerge. These interviews are supplemented by participatory
observation recorded within detailed field notes, and supported by
nearly a decade of sustained involvement and activism by the authors in
the waste management practices of the hospital.

@@

\begin{quote}
Personally, I have never experienced {[}eye discomfort{]} because I get
vaccinated and I am protected including other staff. However, we realize
that the air can badly affect other people and patients who come to this
place.
\end{quote}

@@

Most staff, however, have been unable to find anyway to mitigate the
impacts of the smoke, and only found relief once they reached home at
the end of their shift, as one of the janitorial staff described, ``we
only feel safe when we are home''. Of course, this relief is not
available for the hundreds of patients and caregivers who sleep at QECH
or are unable to leave. Ultimately, most place their hope in the
construction of the new incinerator (which had not yet been activated at
the time of the interviews), and biding their time as construction drags
on; coping as best they can. As one staff member (25/11/2019)
described,``we are just hoping we will start breathing good air soon,
when the new incinerator is opened.''

>>>>

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangerlo@gmail.comm} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[1]{Saloni Vijay%
%
}
 \ead{svijay@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{heiko.heilgendorff@gmail.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10 µm (PM\textsubscript{10})
and less than 2.5 µm (PM\textsubscript{2.5}), expressed as the mass of
PM per volume of air (µg PMx/m\textsuperscript{3} air) were recorded
every 5 minutes at 8 locations across the QECH for 2 months. Qualitative
data collection consisted of 56 interviews with patients, caregivers and
hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators). Our results show that safe air quality
thresholds are consistently exceeded across space and time and that the
most problematic air quality surrounds the shelter for caregivers and
those receiving treatment for HIV/AIDs. Moreover, staff and visitors are
severely impacted by the poor air quality within the space, but feel
powerless to make changes or address complaints. Waste management
interventions are desperately needed lest the patients who arrive at
Queen's leave with more health issues than the ones with which they
arrived.
\end{abstract}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, interior hidden, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Qualitative data collection consisted of 25 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021, spread around the hospital's
various wards and departments. Interviews were semi-structured, and
included questions on hospital waste management practices, perceived
best practices, and workplace health, safety, and risk. The first round
of interviews included specific questions on air quality, while the
second round focused on waste management more broadly. However, the
semi-structured nature of the interviews allowed for the participants to
raise topics of interest, and for unexpected themes to emerge. These
interviews are supplemented by participatory observation recorded within
detailed field notes, and supported by nearly a decade of sustained
involvement and activism by the authors in the waste management
practices of the hospital.

@@

\begin{quote}
Personally, I have never experienced {[}eye discomfort{]} because I get
vaccinated and I am protected including other staff. However, we realise
that the air can badly affect other people and patients who come to this
place.
\end{quote}

@@

Most staff, however, have been unable to find any way to mitigate the
impacts of the smoke, and only found relief once they reached home at
the end of their shift, as one of the janitorial staff described, ``we
only feel safe when we are home''. Of course, this relief is not
available for the hundreds of patients and caregivers who sleep at QECH
or are unable to leave. Ultimately, most place their hope in the
construction of the new incinerator (which had not yet been activated at
the time of the interviews), and biding their time as construction drags
on; coping as best they can. As one staff member (25/11/2019)
described,``we are just hoping we will start breathing good air soon,
when the new incinerator is opened.''
